Regression models for Cylindrical data in Psychology

### 2018-09-24 Manuscript/ManuscriptMarkdownCV.tex
Manuscript update

---
title: "Regression models for Cylindrical data in Psychology"
author: #|
 # | Jolien Cremers^[Corresponding author: j.cremers@uu.nl] $^1$, Helena J.M. Pennings$^{2,3}$,  Christophe Ley $^{4}$
 # | $^1$Department of Methodology and Statistics, Utrecht University
 # | $^2$TNO
  #| $^3$Department of Education, Utrecht University
  #| $^4$Department of Applied Mathematics, Computer Science and Statistics, Ghent University

bibliography: CircularData.bib
csl: apa.csl
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
  html_document: default
  word_document: default
fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage{multirow}
  - \usepackage{appendix}
  - \usepackage{color}
  - \usepackage{hyperref}
  - \usepackage{subcaption}
  - \setlength\parindent{24pt}
  - \usepackage[nodisplayskipstretch]{setspace}\doublespacing
  #- \usepackage{endfloat}
  - \DeclareRobustCommand{\VANDER}[3]{#2} % set up for citation (in .tex %  place \DeclareRobustCommand{\VANDER}[3]{#3} before bibliography)

keywords: bla vla
documentclass: article
pandoc_args: --natbib
abstract: |
  Cylindrical data are multivariate data which consist of a directional,
  in this paper circular, and a linear component. Examples of cylindrical
  data in psychology include human navigation (direction and distance of 
  movement), eye-tracking research (direction and length of saccades) and
  data from an interpersonal circumplex (type and strength of
  interpersonal behavior). In this paper we adapt four models for
  cylindrical data to include a regression of the circular and linear
  component onto a set of covariates. Subsequently, we illustrate how to
  fit these models and interpret their results on a dataset on the
  interpersonal behavior of teachers.
---
```{r inlude = FALSE, echo = FALSE}

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, echo = FALSE}
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
library(plotrix)
library(caret) #cross-validation
```

@@

Cylindrical data are data that consist of a linear variable and a directional
variable. In this paper, the directional variable is circular, meaning that it
consists of a single angle instead of a set of angles. A circular variable is
different from a linear variable in the sense that it is measured on a different
scale. Figure \ref{circline} shows the difference between a circular scale
(right) and a linear scale (left). The most important difference is that on a
circular scale the datapoints 0\(^\circ\) and 360\(^\circ\) are connected and in
fact represent the same number while on a linear scale the two ends, \(-\infty\)
and \(\infty\), are not connected and consequently the values 0\(^\circ\) and
360\(^\circ\) are located on different places on the scale. This difference
requires us to use special statistical methods for circular variables (see e.g.
@fisher1995statistical for an introduction to circular data and
@mardia2000directional, @jammalamadaka2001topics and @ley2017modern for a more
  elaborate overview). For instance, a notion as simple as the sample average
  needs to be re-defined. As is the case for circular data, the analysis of

@@

\indent Cylindrical data occur in several fields of research, such as for
instance in meteorology [@garcia2013exploring], ecology [@garcia2014test] or
marine research [@lagona2015hidden]. However, also in psychology several types
of cylindrical data can be found. For example, in research on human navigation
in the field of cognitive psychology both the distance, a linear variable, and
the direction, a circular variable, of movement are of interest
[@chrastil2017rotational]. In eye-tracking research, saccade data are an example
of cylindrical data, because both the direction (\emph{i.e.}, the circular variable)
and the duration (\emph{i.e.}, the linear variable) of the saccades are of interest
(for a review of eye-tracking research see  @rayner200935th).\newline

@@

psychology are by definition of a cylindrical nature (see Section \ref{Example}
for a more detailed explanation).\newline
\indent In the literature, several methods have been put forward to model the
relation between the linear and circular component of a cylindrical variable.
Some of these are based on regressing the linear component onto the circular
component using the following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is the
linear component and \(\theta\) the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a
different way, e.g. by specifying a multivariate model for several linear and
circular variables and modelling their covariance matrix
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. For
example, @abe2017tractable introduce a cylindrical distribution based on a
Weibull distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their respective
shape and concentration parameters. However, none of the methods that have been
proposed thus far include additional covariates onto which both the circular and
linear component are regressed.\newline
\indent Our aim in this paper is to fill this gap in the literature by adapting
four existing cylindrical models in such a way that they include a regression of
both the linear and circular component of a cylindrical variable onto a set of
covariates. From now on we will therefore refer to the components of the
cylindrical variable as outcome components. Additionally, we will show how a
correct statistical treatment of such cylindrical data can lead to new insights.
We will do this for the teacher data, a dataset from the field of educational
psychology. In the teacher data, apart from modelling the relation between the
linear and circular component of a cylindrical variable we would also like to
predict the two components from a set of covariates in a regression model.

@@

The motivating example for this article comes from the field of educational
psychology and was collected for the studies on classroom climate of
@vanderWant2015role, @Claessens2016side and @pennings2018interpersonal. An
indicator of the quality of the classroom climate is the students' perception of
their teachers' interpersonal behavior. These interpersonal perceptions, both in
educational psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see @horowitz2010handbook for an
overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) [@wubbels2006interpersonal] which is
one such circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and contains items
that load on two interpersonal dimensions: Agency and Communion. Agency refers
to the degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. Different parts of
this space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. This two-dimensional space is called the interpersonal
circle/circumplex (IPC). The IPC is ``a continuous order with no beginning or
end'' [@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering
and the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular
variable.\newline
\indent @Cremers2018Assessing explain the circular nature of the IPC data and
analyze them as such using a circular regression model. The two-dimension scores
Agency and Communion can be converted to a circular score using the two-argument
arctangent function in \eqref{PredVal}, where \(A\) represents a score on the
Agency dimension and \(C\) represents a score on the Communion dimension
\begin{equation}\label{PredVal}
\theta          = \text{atan2}\left(A, \: C\right)  =
\left\{{\begin{array}{lcl}
                                                                       \arctan\left(\frac{A}{C}\right) & \text{if}  \quad&C > 0 \\
\arctan\left(\frac{A}{C}\right) + \pi & \text{if}  \quad& C  <  0  \:\: \&\:\: A \geq 0\\
 \arctan\left(\frac{A}{C}\right) - \pi & \text{if}  \quad&C  <  0 \:\:  \&\:\:A  < 0\\
 +\frac{\pi}{2} & \text{if}  \quad& C  =  0  \:\: \&\:\:A > 0\\
 -\frac{\pi}{2} & \text{if}  \quad& C =  0  \:\: \&\:\:A < 0\\
 \text{undefined} & \text{if} \quad& C =  0   \:\: \&\:\:A = 0.
 \end{array}}
\right.
\end{equation}
The resulting circular variable \(\theta\) can then be modelled and takes values
in the interval $[0, 2\pi)$. However, when two-dimensional data are converted
to the circle we lose some information, namely the length of the two-dimensional
vector \((A, C)^t\), \emph{i.e.}, its Euclidean norm \(\mid\mid (A, C)^t \mid\mid\).
This length represents the strength of the type of interpersonal behavior a
teacher shows towards his/her students and can be considered as the linear
variable in a cylindrical model, allowing us to model a circular variable
\(\theta\) together with the linear variable corresponding to \(\mid\mid (A, C)^t
\mid\mid\). This leads to an improved analysis of interpersonal circumplex data
as we take all information into account. In the next section we introduce
several models that can be used for a more accurate and informative regression
analysis on the teacher data. First however we will provide descriptives for our
data set.

@@

```{r loaddata, return = FALSE, echo = FALSE}
Dat <- read_spss("Data Heleen/mergeddata.sav")

#only select the first measurement occasion
Dat <- subset(Dat, Time == 0)

#sort on class size
Dat <- Dat[order(-Dat$nklas), ]

#Now remove duplicates 
#(because dataframe is sorted on class size the smallest classes are removed)

duplicates <- duplicated(Dat[,c("docnr", "Time")])
Dat <- Dat[!duplicates,]

#center and remove missings for self-efficacy
Dat <- Dat %>% drop_na(Efficacy_CM)
Dat$SEc <- Dat$Efficacy_CM- mean(Dat$Efficacy_CM)
Dat$theta <- atan2(Dat$lAgency, Dat$lcommunion)
Dat$y <- sqrt(Dat$lAgency^2 + Dat$lcommunion^2)

set.seed(30)

#Create holdout and training samples for cross-validation

vec <-  seq(1:nrow(Dat))
flds <- createFolds(vec, k = 10, list = TRUE, returnTrain = FALSE)

for(i in 1:10) {

 #Create name for holdout data 
 nam.h <- paste("hold.", i, sep = "")

 #Get holdout dataset for fold i
 assign(nam.h, Dat[flds[[i]],])

 #Create name for training data
 nam.tr <- paste("train.", i, sep = "")

 #Get training dataset for fold i
 assign(nam.tr, Dat[-flds[[i]],])

}

```

```{r dataplot, cache = TRUE, dev = "tikz",fig.ext="svg", echo = FALSE, results = FALSE}

tikz("Plots/dataplot.tex", standAlone =TRUE, height = 3.5, width = 6, pointsize = 12, engine = "pdftex")

plot(Dat$y, Dat$theta*(180/pi), 
     ylab = "IPC location", xlab = "IPC strength",
     bty = "n", ylim = c(-180, 180))

dev.off()
tools::texi2dvi("Plots/dataplot.tex", pdf=TRUE)

```

```{r descripitves, echo = FALSE, results = FALSE}
#sample size
nrow(Dat)

#summary statistics
summary(Dat$Efficacy_CM)
summary(Dat$y)
summary(as.circular(Dat$theta%%(2*pi)))

sd(Dat$Efficacy_CM)
sd(Dat$y)

```

The teacher data was collected between 2010 and 2015 and contains several
repeated measures on the IPC of 161 teachers. Measurements were obtained using
the QTI and taken in different years and classes. For this paper we only
consider one measurement, the first occasion (2010) and largest class if data
for multiple classes were available. In addition to the score on the IPC, the
circular outcome, and the strength of the score on the IPC, the linear outcome,
a teachers' self-efficacy (\verb|SE|) concerning classroom management is used as
covariate in the analysis. After listwise deletion of missings ($3$ in total,
only for the self-efficacy) we have a sample of 148 teachers. Table
\ref{Tableteacherdescriptives} shows descriptives for the dataset. Here
$\hat{\rho}$ is a sample estimate for the circular concentration where a value
of 0 means that the data is not concentrated at all, \emph{i.e.} spread over the
entire circle, and a value of 1 means that all data is concentrated at a single
point on the circle. Figure \ref{dataplot} is a scatterplot showing the relation
between the linear and circular outcome of the teacher data.

@@

In this section we present four cylindrical models and adapt them such
that they contain predictors for the linear and circular outcomes, \(Y\)
and \(\Theta\). The first two models are based on a construction by
@mastrantonio2015bayesian, while the other models are extensions of the
models from @abe2017tractable and @mastrantonio2018joint.

@@

Following @mastrantonio2015bayesian we consider in this section two models where
the relation between \(\Theta \in [0, 2\pi)\) and \(Y\in (-\infty, + \infty)\)
and $q$ covariates is specified as
\begin{equation}\label{circlinlink}
Y = \gamma_0 + \gamma_{cos}*\cos(\Theta)*R + \gamma_{sin}*\sin(\Theta)*R + \gamma_1*x_1 + \dots + \gamma_q*x_q +  \epsilon,
\end{equation}
\noindent where the random variable \(R\geq0\) will be introduced below, the
error term \(\epsilon \sim N(0, \sigma^2)\) with variance \(\sigma^2>0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are the
intercept and regression coefficients and \(x_1, \dots, x_q\) are the \(q\)
covariates. In both of these models the
conditional distribution of \(Y\) given \(\Theta=\theta\) and \(R = r\) is given
by
\begin{equation}\label{ycondtheta}
f(y \mid \theta, r) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left[-\frac{(y - (\gamma_0 + \gamma_1x_1 + \dots + \gamma_qx_q+c))^{2}}{2\sigma^2}\right],\nonumber
\end{equation}
\noindent where \(c = \begin{bmatrix} r \cos(\theta) \\ r\sin(\theta)
\end{bmatrix}^t \begin{bmatrix} \gamma_{cos} \\ \gamma_{sin} \end{bmatrix}\),
\(r \geq 0\). The linear outcome thus has a normal distribution conditional on
\(\Theta\) and \(R\) and contains already linear covariates \(x_1, \dots, x_q\)
in its location part. For the teacher dataset, the regression equation for the
linear outcome in the CL-PN and CL-GPN model is the following: \[\hat{y}_i =
\gamma_0 + \gamma_{cos}\cos(\theta_i)r_i + \gamma_{sin}\sin(\theta_i)r_i +
\gamma_1\text{SE}_i,\]
\noindent where $\text{SE}_i$ is the self-efficacy score of one individual $i = 1, \dots, n$ where $n$ is the sample size.\newline
\indent For the circular outcome we assume either a projected normal (PN) or a general
projected normal (GPN) distribution. These distributions arise from the radial
projection of a distribution defined on the plane onto the circle. The relation
between a bivariate vector \(\boldsymbol{S}\) in the plane and the circular
outcome \(\Theta\) is defined as follows
\begin{equation}\label{projection}
\boldsymbol{S} = \begin{bmatrix} S^{I} \\ S^{II} \end{bmatrix} = R\boldsymbol{u} = \begin{bmatrix} R \cos (\Theta) \\  R\sin (\Theta) \end{bmatrix},
\end{equation}
\noindent where \(R = \mid\mid \boldsymbol{S} \mid\mid\), the Euclidean norm of
the bivariate vector \(\boldsymbol{S}\). In the PN distribution we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{I})\) and in the GPN we
assume \(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})\) where
$\boldsymbol{\mu} \in \mathbb{R}^2$, \(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2 & \rho\\ \rho & 1
\end{bmatrix}\), \(\rho \in (-\infty, +\infty)\) and \(\tau^2 \geq 0\) (as in
@hernandez2016general). This leads to the
circular-linear PN (CL-PN) and circular-linear GPN (CL-GPN) distributions. We
will now detail how we modify both cylindrical distributions to also incorporate
covariates for the circular part.

@@

Following @wang2012directional and @hernandez2016general the joint density of
\(R\) and \(\Theta\) for the GPN distribution equals
\begin{equation}\label{gpnreg}
f(\theta, r \mid \boldsymbol{\mu}, \boldsymbol{\Sigma}) = \frac{r}{2\pi\tau} \exp\left[ -\frac{(r\boldsymbol{u}-\boldsymbol{\mu})^{t}\boldsymbol{\Sigma}^{-1}(r\boldsymbol{u}-\boldsymbol{\mu})}{2}\right],
\end{equation}
\noindent where we recall that \(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 +
\rho^2 & \rho\\ \rho & 1 \end{bmatrix}\). In a regression setup the outcomes
\(\theta_i\) and \(r_i\) for each individual are generated independently from
\eqref{gpnreg}. The mean vector \(\boldsymbol{\mu}_i \in \mathbb{R}^2\) is
defined in the same way via covariates as for the modified CL-PN distribution.
Note in contrast with the CL-PN model where \(\boldsymbol{z}_i^I\) and
\(\boldsymbol{z}_i^{II}\) are allowed to differ we do need to have the same
predictors for both components of \(\boldsymbol{\mu}_i\) in the CL-GPN model.
This is due to the fact that the variance-covariance matrix
\(\boldsymbol{\Sigma}\) is no longer identity. For the teacher dataset, the
regression equation for the circular outcome in the CL-GPN model is the same as
in the CL-PN model.

@@

```{r FitCLPN, cache = TRUE, results = FALSE, echo = FALSE, eval = TRUE}

source("R-code/Posterior Sampling CL-PN.R")

its <- 20000
set.seed(101)

res_CLPN <- list()

for(i in 1:10){

  train <- get(paste("train.", i, sep = ""))
  hold <- get(paste("hold.", i, sep = ""))

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Regression models for Cylindrical data in Psychology},
            pdfkeywords={bla vla},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Regression models for Cylindrical data in Psychology}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\usepackage{multirow}
\usepackage{appendix}
\usepackage{color}
\usepackage{hyperref}
\usepackage{subcaption}
\setlength\parindent{24pt}
\usepackage[nodisplayskipstretch]{setspace}\doublespacing
\DeclareRobustCommand{\VANDER}[3]{#2} % set up for citation (in .tex %  place \DeclareRobustCommand{\VANDER}[3]{#3} before bibliography)

\begin{document}
\maketitle
\begin{abstract}
Cylindrical data are multivariate data which consist of a directional,
in this paper circular, and a linear component. Examples of cylindrical
data in psychology include human navigation (direction and distance of
movement), eye-tracking research (direction and length of saccades) and
data from an interpersonal circumplex (type and strength of
interpersonal behavior). In this paper we adapt four models for
cylindrical data to include a regression of the circular and linear
component onto a set of covariates. Subsequently, we illustrate how to
fit these models and interpret their results on a dataset on the
interpersonal behavior of teachers.
\end{abstract}

@@

Cylindrical data are data that consist of a linear variable and a
directional variable. In this paper, the directional variable is
circular, meaning that it consists of a single angle instead of a set of
angles. A circular variable is different from a linear variable in the
sense that it is measured on a different scale. Figure \ref{circline}
shows the difference between a circular scale (right) and a linear scale
(left). The most important difference is that on a circular scale the
datapoints 0\(^\circ\) and 360\(^\circ\) are connected and in fact
represent the same number while on a linear scale the two ends,
\(-\infty\) and \(\infty\), are not connected and consequently the
values 0\(^\circ\) and 360\(^\circ\) are located on different places on
the scale. This difference requires us to use special statistical
methods for circular variables (see e.g. Fisher (1995) for an
introduction to circular data and Mardia \& Jupp (2000), Jammalamadaka
\& Sengupta (2001) and Ley \& Verdebout (2017) for a more elaborate
overview). For instance, a notion as simple as the sample average needs
to be re-defined. As is the case for circular data, the analysis of

@@

\indent Cylindrical data occur in several fields of research, such as
for instance in meteorology (García-Portugués, Crujeiras, \&
González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras,
González-Manteiga, \& Pereira, 2014) or marine research (Lagona, Picone,
Maruotti, \& Cosoli, 2015). However, also in psychology several types of
cylindrical data can be found. For example, in research on human
navigation in the field of cognitive psychology both the distance, a
linear variable, and the direction, a circular variable, of movement are
of interest (Chrastil \& Warren, 2017). In eye-tracking research,
saccade data are an example of cylindrical data, because both the
direction (\emph{i.e.}, the circular variable) and the duration
(\emph{i.e.}, the linear variable) of the saccades are of interest (for
a review of eye-tracking research see Rayner (2009)).\newline

@@

psychology are by definition of a cylindrical nature (see Section
\ref{Example} for a more detailed explanation).\newline
\indent In the literature, several methods have been put forward to
model the relation between the linear and circular component of a
cylindrical variable. Some of these are based on regressing the linear
component onto the circular component using the following type of
relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is
the linear component and \(\theta\) the circular component (Johnson \&
Wehrly, 1978; Mardia \& Sutton, 1978; Mastrantonio, Maruotti, \&
Jona-Lasinio, 2015). Others model the relation in a different way,
e.g.~by specifying a multivariate model for several linear and circular
variables and modelling their covariance matrix (Mastrantonio, 2018) or
by proposing a joint cylindrical distribution. For example, Abe \& Ley
(2017) introduce a cylindrical distribution based on a Weibull
distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their
respective shape and concentration parameters. However, none of the
methods that have been proposed thus far include additional covariates
onto which both the circular and linear component are regressed.\newline
\indent Our aim in this paper is to fill this gap in the literature by
adapting four existing cylindrical models in such a way that they
include a regression of both the linear and circular component of a
cylindrical variable onto a set of covariates. From now on we will
therefore refer to the components of the cylindrical variable as outcome
components. Additionally, we will show how a correct statistical
treatment of such cylindrical data can lead to new insights. We will do
this for the teacher data, a dataset from the field of educational
psychology. In the teacher data, apart from modelling the relation
between the linear and circular component of a cylindrical variable we
would also like to predict the two components from a set of covariates
in a regression model.

@@

The motivating example for this article comes from the field of
educational psychology and was collected for the studies on classroom
climate of \VANDER{Want}{Van der}{van der} Want (2015), Claessens (2016)
and Pennings et al. (2018). An indicator of the quality of the classroom
climate is the students' perception of their teachers' interpersonal
behavior. These interpersonal perceptions, both in educational
psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see Horowitz \& Strack (2011)
for an overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) (Wubbels, Brekelmans, Brok,
\& Tartwijk, 2006) which is one such circumplex measurement instrument.
The QTI is designed to measure student perceptions of their teachers'
interpersonal behavior and contains items that load on two interpersonal
dimensions: Agency and Communion. Agency refers to the degree of power
or control a teacher exerts in interaction with his/her students.
Communion refers to the degree of friendliness or affiliation a teacher
conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by
Agency (vertical) and Communion (horizontal), see Figure \ref{QTI}.
Different parts of this space are characterized by different teacher
behavior, e.g. `helpful' or `uncertain'. This two-dimensional space is
called the interpersonal circle/circumplex (IPC). The IPC is ``a
continuous order with no beginning or end'' (Gurtman, 2009, p. 2). We
call such ordering a circumplex ordering and the IPC is therefore often
called the interpersonal circumplex. The ordering also implies that
scores on the IPC could be viewed as a circular variable.\newline
\indent Cremers, Mainhard, \& Klugkist (2018a) explain the circular
nature of the IPC data and analyze them as such using a circular
regression model. The two-dimension scores Agency and Communion can be
converted to a circular score using the two-argument arctangent function
in \eqref{PredVal}, where \(A\) represents a score on the Agency
dimension and \(C\) represents a score on the Communion dimension

\begin{equation}\label{PredVal}
\theta          = \text{atan2}\left(A, \: C\right)  =
\left\{{\begin{array}{lcl}
                                                                       \arctan\left(\frac{A}{C}\right) & \text{if}  \quad&C > 0 \\
\arctan\left(\frac{A}{C}\right) + \pi & \text{if}  \quad& C  <  0  \:\: \&\:\: A \geq 0\\
 \arctan\left(\frac{A}{C}\right) - \pi & \text{if}  \quad&C  <  0 \:\:  \&\:\:A  < 0\\
 +\frac{\pi}{2} & \text{if}  \quad& C  =  0  \:\: \&\:\:A > 0\\
 -\frac{\pi}{2} & \text{if}  \quad& C =  0  \:\: \&\:\:A < 0\\
 \text{undefined} & \text{if} \quad& C =  0   \:\: \&\:\:A = 0.
 \end{array}}
\right.
\end{equation}

The resulting circular variable \(\theta\) can then be modelled and
takes values in the interval \([0, 2\pi)\). However, when
two-dimensional data are converted to the circle we lose some
information, namely the length of the two-dimensional vector
\((A, C)^t\), \emph{i.e.}, its Euclidean norm
\(\mid\mid (A, C)^t \mid\mid\). This length represents the strength of
the type of interpersonal behavior a teacher shows towards his/her
students and can be considered as the linear variable in a cylindrical
model, allowing us to model a circular variable \(\theta\) together with
the linear variable corresponding to \(\mid\mid (A, C)^t \mid\mid\).
This leads to an improved analysis of interpersonal circumplex data as
we take all information into account. In the next section we introduce
several models that can be used for a more accurate and informative
regression analysis on the teacher data. First however we will provide
descriptives for our data set.

@@

The teacher data was collected between 2010 and 2015 and contains
several repeated measures on the IPC of 161 teachers. Measurements were
obtained using the QTI and taken in different years and classes. For
this paper we only consider one measurement, the first occasion (2010)
and largest class if data for multiple classes were available. In
addition to the score on the IPC, the circular outcome, and the strength
of the score on the IPC, the linear outcome, a teachers' self-efficacy
(\verb|SE|) concerning classroom management is used as covariate in the
analysis. After listwise deletion of missings (\(3\) in total, only for
the self-efficacy) we have a sample of 148 teachers. Table
\ref{Tableteacherdescriptives} shows descriptives for the dataset. Here
\(\hat{\rho}\) is a sample estimate for the circular concentration where
a value of 0 means that the data is not concentrated at all, \emph{i.e.}
spread over the entire circle, and a value of 1 means that all data is
concentrated at a single point on the circle. Figure \ref{dataplot} is a
scatterplot showing the relation between the linear and circular outcome
of the teacher data.

@@

In this section we present four cylindrical models and adapt them such
that they contain predictors for the linear and circular outcomes, \(Y\)
and \(\Theta\). The first two models are based on a construction by
Mastrantonio et al. (2015), while the other models are extensions of the
models from Abe \& Ley (2017) and Mastrantonio (2018).

@@

Following Mastrantonio et al. (2015) we consider in this section two
models where the relation between \(\Theta \in [0, 2\pi)\) and
\(Y\in (-\infty, + \infty)\) and \(q\) covariates is specified as

\begin{equation}\label{circlinlink}
Y = \gamma_0 + \gamma_{cos}*\cos(\Theta)*R + \gamma_{sin}*\sin(\Theta)*R + \gamma_1*x_1 + \dots + \gamma_q*x_q +  \epsilon,
\end{equation}

\noindent where the random variable \(R\geq0\) will be introduced below,
the error term \(\epsilon \sim N(0, \sigma^2)\) with variance
\(\sigma^2>0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are
the intercept and regression coefficients and \(x_1, \dots, x_q\) are
the \(q\) covariates. In both of these models the conditional
distribution of \(Y\) given \(\Theta=\theta\) and \(R = r\) is given by

\begin{equation}\label{ycondtheta}
f(y \mid \theta, r) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left[-\frac{(y - (\gamma_0 + \gamma_1x_1 + \dots + \gamma_qx_q+c))^{2}}{2\sigma^2}\right],\nonumber
\end{equation}

\noindent where
\(c = \begin{bmatrix} r \cos(\theta) \\ r\sin(\theta) \end{bmatrix}^t \begin{bmatrix} \gamma_{cos} \\ \gamma_{sin} \end{bmatrix}\),
\(r \geq 0\). The linear outcome thus has a normal distribution
conditional on \(\Theta\) and \(R\) and contains already linear
covariates \(x_1, \dots, x_q\) in its location part. For the teacher
dataset, the regression equation for the linear outcome in the CL-PN and
CL-GPN model is the following: \[\hat{y}_i =
\gamma_0 + \gamma_{cos}\cos(\theta_i)r_i + \gamma_{sin}\sin(\theta_i)r_i +
\gamma_1\text{SE}_i,\] \noindent where \(\text{SE}_i\) is the
self-efficacy score of one individual \(i = 1, \dots, n\) where \(n\) is
the sample size.\newline
\indent For the circular outcome we assume either a projected normal
(PN) or a general projected normal (GPN) distribution. These
distributions arise from the radial projection of a distribution defined
on the plane onto the circle. The relation between a bivariate vector
\(\boldsymbol{S}\) in the plane and the circular outcome \(\Theta\) is
defined as follows

\begin{equation}\label{projection}
\boldsymbol{S} = \begin{bmatrix} S^{I} \\ S^{II} \end{bmatrix} = R\boldsymbol{u} = \begin{bmatrix} R \cos (\Theta) \\  R\sin (\Theta) \end{bmatrix},
\end{equation}

\noindent where \(R = \mid\mid \boldsymbol{S} \mid\mid\), the Euclidean
norm of the bivariate vector \(\boldsymbol{S}\). In the PN distribution
we assume \(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{I})\)
and in the GPN we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})\) where
\(\boldsymbol{\mu} \in \mathbb{R}^2\),
\(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2 & \rho\\ \rho & 1 \end{bmatrix}\),
\(\rho \in (-\infty, +\infty)\) and \(\tau^2 \geq 0\) (as in
Hernandez-Stumpfhauser, Breidt, \& Woerd (2016)). This leads to the
circular-linear PN (CL-PN) and circular-linear GPN (CL-GPN)
distributions. We will now detail how we modify both cylindrical
distributions to also incorporate covariates for the circular part.

@@

Following Wang \& Gelfand (2013) and Hernandez-Stumpfhauser et al.
(2016) the joint density of \(R\) and \(\Theta\) for the GPN
distribution equals

\begin{equation}\label{gpnreg}
f(\theta, r \mid \boldsymbol{\mu}, \boldsymbol{\Sigma}) = \frac{r}{2\pi\tau} \exp\left[ -\frac{(r\boldsymbol{u}-\boldsymbol{\mu})^{t}\boldsymbol{\Sigma}^{-1}(r\boldsymbol{u}-\boldsymbol{\mu})}{2}\right],
\end{equation}

\noindent where we recall that
\(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2 & \rho\\ \rho & 1 \end{bmatrix}\).
In a regression setup the outcomes \(\theta_i\) and \(r_i\) for each
individual are generated independently from \eqref{gpnreg}. The mean
vector \(\boldsymbol{\mu}_i \in \mathbb{R}^2\) is defined in the same
way via covariates as for the modified CL-PN distribution. Note in
contrast with the CL-PN model where \(\boldsymbol{z}_i^I\) and
\(\boldsymbol{z}_i^{II}\) are allowed to differ we do need to have the
same predictors for both components of \(\boldsymbol{\mu}_i\) in the
CL-GPN model. This is due to the fact that the variance-covariance
matrix \(\boldsymbol{\Sigma}\) is no longer identity. For the teacher
dataset, the regression equation for the circular outcome in the CL-GPN
model is the same as in the CL-PN model.

@@
